---
title: "Process Data For Figures"
output: html_document
date: "2026-02-02"
---

# NOTE: All comments starting with `###` are internal notes and should be adressed and removed before pushing to repo. 

# Description
- Given that a lot of the data used by the figures is shared, this script pre-processes it across the figures in order to make them easier to interpret
- Also helpful as many of the data files are large, so pre-processing them once saves time and computational resources
- The output will be a set of .RDS files for use across the various figures, with as much pre-processing as possible
- This version uses the updated data from the breadth and bias pre-filtering

# Load Setup and Config
```{r setup, include=FALSE}
library(install.load)

install_load("here")

source(here("00_R_configs", "config.R"))

source(here("00_R_configs", "00_setup.R"))

source(here("00_R_configs", "00_functions.R"))
``` 

# Load Input Data
```{r}
# ---- Load Phylogenetic Tree ----
phylo_tree_30k <- read.tree(data_here("00_data", "parsnp.tree"))

# ---- Load consenTrait Data ----
consenTrait_data <- readRDS(data_here("00_data", "consentrait_depth_results.rds")) 

# ---- Load ANI Data ----
ani_data <- data_here("/00_data/fastANI_output", "ANI_clusters") %>%
  list.files(pattern = "\\.csv$", full.names = TRUE) %>%
  purrr::map(~{
    df <- read_csv(.x)
    cluster <- str_extract(basename(.x), "\\d+\\.\\d+")
    colnames(df)[2] <- cluster
    df
  }) %>%
  reduce(full_join, by = colnames(.[[1]])[1])

# ---- Load Metadata ----
metadata <- read.csv(data_here("00_data", "blockedModelInputMetadata.csv"))

genomes_in_train <- metadata %>%
  filter(split == "Train") %>%
  pull(Genome_ID) %>%
  unique()

# ---- Load full PA 30K genome metadata ----
pa_30k_full <- read.delim(data_here("00_data", "PA_30K_GenomeMetadata.tsv")) %>%
  mutate(Assembly.Accession = gsub("\\.", "_", Assembly.Accession)) %>%
  rename(Genome_ID = Assembly.Accession)

# ---- Load Blocked Cross Performance Data ----
blocked_performance_data <- data_here("00_data", "Blocked_Cross_Results") %>%
  list.files(pattern = "_holdout_stats_newPanaroo\\.csv$", full.names = TRUE) %>%
  set_names(~str_extract(basename(.), "^[a-z]+") %>% str_to_title()) %>%
  purrr::map(read.csv) 
  # Loads everything in the folder, needs to be filtered downstream

# ---- Load Final Model Performance Data ----
final_performance_data <- read.csv(data_here("00_data/model_output/full_model_results", "477_filtered_test.csv"))

# ---- Load Processed SHAP Data ----
final_iterative_genes <- readRDS(data_here("00_data/model_output", "shap_genes_annotated_blocked_model.rds"))

# ---- Load Final Model ROC Data ----
fpr_data <- read.csv(data_here("00_data/model_output/roc", "477model_fpr.csv"))

tpr_data <- read.csv(data_here("00_data/model_output/roc", "477model_tpr.csv"))

raw_probs <- read.csv(data_here("00_data/model_output/roc", "477model_raw_probs.csv"))

# ---- Load Other Processed Data ----
genes_of_interest <- final_iterative_genes %>% 
    bind_rows() %>%
    select(gene) %>%
    distinct() %>%
    unlist() %>%
    as.character()

###############################################################################
###############################################################################
###############################################################################
###############################################################################
###############################################################################
########################## STOPPED HERE ####################################### 
# (where is gene_presence_absence_summary generated?) (as of 7/1 still need
# to check)
# As of 7/1 added comments for changes downstream, need to go through
###############################################################################
###############################################################################
###############################################################################
###############################################################################
###############################################################################

gene_presence_absence_summary <- read.csv(
  data_here("00_data", "model_input", "gene_presence_absence_summary.csv")
) %>% as_tibble()

# ---- Load Model Input Data ----
model_input_filtered <- fread(
    data_here("00_data", "model_input", "ModelInput_NoDuplicates.csv"),
    select = c("Genome_ID", (genes_of_interest))
)

# ---- Load Iris Gene Data ----
pa_gene_ids <- read.csv(
  data_here("00_data", "PAO1_blast_result_new.csv"))

pa_gene_ids %>% colnames()

# DOES NOT include genes: group_478266 and group_643509, which don't match to PAO1 gene

# ---- Load All Gene Funct/No_Funct Data ----
iris_output_dir <- data_here("00_data", "outputs_for_graphing_chronic")

# ---- Load "Final" Gene Function Data ----
iris_output_dir_v3 <- data_here("00_data", "CCS-topgenesv3")

# ---- Load Experimental Data ----
Experimental_Metadata <- read.csv2(data_here("00_data", "Working_Experimental_Metadata_WithID.csv"))

cladeData <- read.csv2(data_here("00_data", "Key Strain List (Working Library).csv"), sep = ",") %>% 
  select(Clade, Library_ID)

# ---- Load and Root Experimental Tree ----
Experimental_Tree <- read.tree(data_here("00_data", "Experimental_iqtree_1000.contree")) %>%
  root(outgroup = "PA7.fasta", resolve.root = TRUE)
```

# ============================================
# Process Metadata
# ============================================

## Define "Final" Niche Mapping Used Across Figures
```{r}
# ---- Define Niche Mapping ----
niche_map <- c(
  "Bronchiectasis" = "Bronchiectasis",
  "CF" = "Adult CF",
  "early.CF" = "Pediatric CF",
  "Early CF" = "Pediatric CF",
  "Early.CF" = "Pediatric CF",
  "Urinary" = "Urinary",
  "Gastrointestinal" = "Gastrointestinal",
  "Wound" = "Wound",
  "Blood" = "Blood",
  "Aquatic" = "Aquatic",
  "Ocean" = "Aquatic",
  "Hospital Environment" = "Hospital",
  "Hospital.Environment" = "Hospital",
  "Hospital.Enviornment" = "Hospital",
  "Rectal/Feces" = "Fecal",
  "Rectal-Feces" = "Fecal",
  "Rectal.Feces" = "Fecal"
)

remap_niches <- function(niche_vector) {
  recode(niche_vector, !!!niche_map, .default = NA_character_)
}

# ---- Apply to consenTrait Data ----
consenTrait_data <- consenTrait_data %>%
  mutate(Niche = remap_niches(Niche))

# ---- Apply to Metadata ----
metadata_mapped <- metadata_unmapped %>%
  mutate(Niche = remap_niches(Niche)) %>%
  filter(!is.na(Niche)) %>%
  filter(Clade %in% "CladeA")

# ---- Apply to SHAP data (rename list elements) ----
final_iterative_genes %>% select(Env) %>% distinct() 
# Already good here

# ---- Helper Function: Apply Niche Mapping to Label Strings ----
remap_labels <- function(label_string, niche_map) {
  result <- label_string
  for (old_name in names(niche_map)) {
    new_name <- niche_map[[old_name]]
    old_pattern <- paste0("'", old_name, "'")
    new_replacement <- paste0("'", new_name, "'")
    result <- gsub(old_pattern, new_replacement, result, fixed = TRUE)
  }
  return(result)
}

# ---- Apply to Performance Data (rename in True/Predicted Labels) ----
blocked_performance_data <- lapply(blocked_performance_data, function(df) {
  df %>%
    mutate(
      True.Labels = remap_labels(True.Labels, niche_map),
      Predicted.Labels = remap_labels(Predicted.Labels, niche_map)
    )
})

eleventhRound <- blocked_performance_data$Eleventh

final_performance_data <- final_performance_data %>%
  mutate(
    True.Labels = remap_labels(True.Labels, niche_map),
    Predicted.Labels = remap_labels(Predicted.Labels, niche_map)
  )

# ---- Apply to ROC Data ----
fpr_data_long <- fpr_data %>%
  pivot_longer(
    cols = -X,
    names_to = "Env",
    values_to = "FPR"
  )

tpr_data_long <- tpr_data %>%
  pivot_longer(
    cols = -X,
    names_to = "Env",
    values_to = "TPR"
  )

# Merge FPR and TPR on X and Env
roc_data <- fpr_data_long %>%
  left_join(tpr_data_long, by = c("X", "Env")) %>%
  mutate(Env = remap_niches(Env))

saveRDS(roc_data, 
        file = data_here("00_data", "Figure_Processed_Data", "477_roc_data_processed.rds"))

# Rename columns in raw_probs and apply niche mapping

probs_long <- raw_probs %>%
  pivot_longer(
    cols = -Genome_IDs,
    names_to = "Env",
    values_to = "Probability"
  ) %>%
  mutate(Env = remap_niches(Env))

saveRDS(probs_long, 
        file = data_here("00_data", "Figure_Processed_Data", "477_probs_processed.rds"))

focal_env_counts <- probs_long %>%
  group_by(Genome_IDs) %>%
  mutate(
    Focal_Env = Env[Probability == max(Probability)][1]
  ) %>%
  ungroup() %>%
  distinct(Genome_IDs, Focal_Env) %>%
  group_by(Focal_Env) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))
```

## Filter Metadata to Clade A and Map to Valid Niches
```{r}
saveRDS(metadata_mapped, 
        file = data_here("00_data", "Figure_Processed_Data",  "metadata_mapped.rds"))

metadata_filtered <- metadata_mapped %>% 
  filter(Clade == "CladeA", !is.na(Niche)) %>%
  mutate(Genome_ID = as.character(Genome_ID)) %>%
  filter(Genome_ID %in% genomes_in_train)

  # ---- Print Stats for Niche Counts ----
niche_counts <- metadata_filtered %>%
  group_by(Niche) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  arrange(desc(Count))

print(niche_counts)

saveRDS(metadata_filtered, 
        file = data_here("00_data", "Figure_Processed_Data", "Metadata_Filtered.rds"))
```

# ============================================
# Figure 1
# ============================================

## Process Phylogenetic Tree Data
```{r}
#  ---- Clean Tip Labels ----
phylo_tree_30k$tip.label <- gsub(".fna$", "", phylo_tree_30k$tip.label)

# ---- Add Outgroup ----
outgroup_tip <- "PA7_GCA_000017205.1_ASM1720v1_genomic"

if (!(outgroup_tip %in% metadata_filtered$Genome_ID)) {
  outgroup_row <- data.frame(
    Genome_ID = outgroup_tip, 
    Clade = "Outgroup", 
    Niche = "Outgroup", 
    Genomovar_Cluster = "Outgroup", 
    Clone_Cluster = "Outgroup"
  )
  metadata_filtered <- rbind(metadata_filtered, outgroup_row)
}

# ---- Filter and Root Tree ----
phylo_filtered_rooted <- phylo_tree_30k %>%
  ape::keep.tip(phylo_tree_30k$tip.label[phylo_tree_30k$tip.label %in% metadata_filtered$Genome_ID]) %>%
  root(outgroup = outgroup_tip, resolve.root = TRUE)

# ---- Join Metadata to Tree ----

phylo_filtered_rooted_treedata <- as.treedata(phylo_filtered_rooted)

metadata_for_join <- metadata_filtered %>%
  mutate(label = Genome_ID) %>%
  select(label, Niche)

phylo_filtered_rooted_treedata <- full_join(
  phylo_filtered_rooted_treedata, 
  metadata_for_join, 
  by = "label"
)

# ---- Save Processed Phylogenetic Tree Data ----
saveRDS(phylo_filtered_rooted_treedata, 
        file = data_here("00_data", "Figure_Processed_Data", "phylo_filtered_rooted_treedata.RDS"))
```

## Process consenTrait Data
```{r}
# ---- Extract Tree and Trait Data ----
tree <- phylo_filtered_rooted_treedata@phylo
metadata <- as_tibble(phylo_filtered_rooted_treedata) %>%
  filter(!is.na(label))  # Keep only tips (remove internal nodes)

# Reorder metadata to match tree tip labels EXACTLY
metadata <- metadata[match(tree$tip.label, metadata$label), ]

niche_order <- c(
  "Adult CF", "Pediatric CF", "Bronchiectasis", 
  "Blood", "Urinary", "Fecal", "Wound",
  "Hospital",
  "Aquatic")

# ==============================================================================
# CONSTRAINT DEPTH ANALYSIS
# ==============================================================================

# ---- Run Analysis in Parallel ----
message("Starting constraint depth calculations...")
consentrait_results <- mclapply(
  niche_order, 
  calculate_consentrait_depth, 
  tree = tree, 
  metadata = metadata, 
  mc.cores = 10
)

consentrait_results <- bind_rows(consentrait_results) %>% 
  filter(!is.na(Mean_depth)) %>%
  mutate(
    P_adj = p.adjust(P_value, method = "BH"),
    Seq_dissimilarity = 2 * Mean_depth,
    Seq_similarity = 1 - Seq_dissimilarity,
    Seq_similarity_pct = 100 * Seq_similarity, 
    Seq_similarity_pct_Max = (1 - (2 * Max_depth)) * 100
  )

# ---- Display Results ----
cat("\n=== CONSTRAINT DEPTH RESULTS ===\n")
print(consentrait_results)

consentrait_results %>% select(Niche, P_value, Seq_similarity_pct, Seq_similarity_pct_Max)

saveRDS(consenTrait_data, 
        file = data_here("00_data", "Figure_Processed_Data", "consenTrait_data.RDS"))
```

## Process ANI Data
```{r}
# Already processed, just re-saving for consistency

saveRDS(ani_data, 
        file = data_here("00_data", "Figure_Processed_Data", "ani_data.RDS"))
```

# ============================================
# Figure 2
# ============================================

## Process Blocked Confusion Matrix Data
```{r}
# ---- Confusion Matrix from Blocked Models ----
make_confusion_from_blocked <- function(df_blocked) {
  # Keep only blocked models
  df_blocked <- df_blocked %>%
    dplyr::filter(stringr::str_detect(model, "_isolated$"))
  
  # Expand each row into per-isolate true/pred pairs
  pairs_df <- purrr::map_dfr(seq_len(nrow(df_blocked)), function(i) {
    true_envs <- stringr::str_extract_all(df_blocked$True.Labels[i], "'([^']+)'")[[1]] %>%
      gsub("'", "", .)
    pred_envs <- stringr::str_extract_all(df_blocked$Predicted.Labels[i], "'([^']+)'")[[1]] %>%
      gsub("'", "", .)
    
    # Safety check: lengths should match
    if (length(true_envs) != length(pred_envs)) {
      stop(paste0(
        "Row ", i, " has mismatched label counts: ",
        length(true_envs), " true vs ", length(pred_envs), " pred"
      ))
    }
    
    tibble::tibble(
      True = true_envs,
      Pred = pred_envs
    )
  })
  
  # Use unified processing function
  process_confusion_matrix(pairs_df)
}

cm_blocked <- make_confusion_from_blocked(eleventhRound)

# ---- Define environment order ----
env_labels <- c(
  "Aquatic", "Hospital",
  "Adult CF", "Pediatric CF", "Bronchiectasis", "Blood", "Urinary", "Fecal", "Wound"
)

accuracy_df <- calculate_accuracy_metrics(cm_blocked, env_labels)

# ---- Save Processed Blocked Confusion Matrix Data ----
saveRDS(list(
  confusion_matrix = cm_blocked,
  accuracy_metrics = accuracy_df
),
file = data_here("00_data", "Figure_Processed_Data", "blocked_confusion_matrix_data.RDS"))
```

## Process "Final" Model Confusion Matrix Data 
```{r}
# ---- Confusion Matrix from Single Model Row ----
make_confusion_from_final <- function(df_row) {
  # Validation
  if (!is.data.frame(df_row)) {
    stop("df_row must be a data frame")
  }
  if (nrow(df_row) == 0) {
    stop("df_row has 0 rows; please provide a pre-filtered data frame with one model row")
  }
  if (nrow(df_row) > 1) {
    stop("df_row has more than one row; please pre-filter to a single model row")
  }
  
  # Extract labels from the single row
  true_envs <- stringr::str_extract_all(df_row$True.Labels[1], "'([^']+)'")[[1]] %>%
    gsub("'", "", .)
  pred_envs <- stringr::str_extract_all(df_row$Predicted.Labels[1], "'([^']+)'")[[1]] %>%
    gsub("'", "", .)
  
  # Safety check: lengths should match
  if (length(true_envs) != length(pred_envs)) {
    stop(paste0(
      "Selected model row has mismatched label counts: ",
      length(true_envs), " true vs ", length(pred_envs), " pred"
    ))
  }
  
  pairs_df <- tibble::tibble(
    True = true_envs,
    Pred = pred_envs
  )
  
  # Use unified processing function
  process_confusion_matrix(pairs_df)
}

cm_final <- make_confusion_from_final(final_performance_data)

accuracy_df_final <- calculate_accuracy_metrics(cm_final, env_labels)

final_performance_stats <- final_performance_data %>% select(-True.Labels, -Predicted.Labels)

# ---- Save Processed Final Confusion Matrix Data ----
saveRDS(list(
  stats = final_performance_stats,
  confusion_matrix = cm_final,
  accuracy_metrics = accuracy_df_final
),
file = data_here("00_data", "Figure_Processed_Data", "Final_Model_Performance_Data_Processed.RDS"))
```


# ============================================
# Figure 3
# ============================================

## Categorize Genes in SHAP Data
```{r}
# ---- Merge SHAP Data Across Niches ----
final_iterative_genes

# ---- Check that all clusters have the same annotation ----
annotationConsensus <- final_iterative_genes %>%
    bind_rows() %>%
    group_by(aa_cluster) %>%
    mutate(Annotation_Consensus = n_distinct(Annotation_Inferred) == 1) %>%
    ungroup() %>%
    select(aa_cluster, Annotation_Inferred, Annotation_Consensus) %>%
    filter(!Annotation_Consensus)

annotationConsensus # Should be 0 rows

# ---- Map Functional Categories and Aggregate Counts ----
categorize_all_genes <- function(df, anno_col = "Annotation_Inferred") {
  out_list <- lapply(df[[anno_col]], categorize_two_tier_per_part) 
  # See 00_functions.R for function definition
  out_df <- bind_rows(out_list, .id = "RowID") %>%
    mutate(RowID = as.integer(RowID))
  
  df_out <- df %>%
    mutate(RowID = row_number()) %>%
    left_join(out_df, by = "RowID") %>%
    select(-RowID)
  
  return(df_out)
}

flat_df_annotated <- categorize_all_genes(final_iterative_genes)

flat_df_annotated_gene_names <- flat_df_annotated %>%
  group_by(aa_cluster) %>%
  mutate(
    # Find the Non.unique.Gene.name of the longest sequence (aa_length) in the cluster
    cluster_gene_name = {
      longest_idx <- which.max(aa_length)
      non_blank_name <- Non.unique.Gene.name[longest_idx]
      if (length(longest_idx) > 0 && !is.na(non_blank_name) && trimws(non_blank_name) != "") {
        non_blank_name
      } else {
        NA_character_
      }
    },
    # Use cluster gene name if current row's gene name is NA or empty
    Non.unique.Gene.name = case_when(
      is.na(Non.unique.Gene.name) | trimws(Non.unique.Gene.name) == "" ~ cluster_gene_name,
      TRUE ~ Non.unique.Gene.name
    )
  ) %>%
  ungroup() %>%
  select(-cluster_gene_name)

# ---- Save Annotated SHAP Data ----
saveRDS(flat_df_annotated, 
        file = data_here("00_data", "Figure_Processed_Data", "flat_df_annotated_SHAP_data.RDS"))

saveRDS(flat_df_annotated_gene_names, 
        file = data_here("00_data", "Figure_Processed_Data", "flat_df_annotated_gene_names_SHAP_data.RDS"))
```

## Export Annotated SHAP Data for genes (per ENV)
```{r}
unique_envs <- flat_df_annotated_gene_names$Env %>% unique()

for (Envs in unique_envs) {
  niche_data <- flat_df_annotated_gene_names %>%
    filter(Env == Envs) %>%
    arrange(desc(mean_abs_shap))
  filename <- data_here("00_data", 
    "Figure_Processed_Data", 
    "Processed_Final_SHAP_Per_ENV", 
    "Bias_Filtered_Features_UpdatedClusters_Annotated", 
    paste0("flat_df_annotated_", Envs, ".csv"))
  write.csv(niche_data, filename, row.names = FALSE)
}
```

## Process Breadth Data
```{r}
### I think we don't use this anymore. DOUBLE CHECK.

final_iterative_genes_breadth <- final_iterative_genes %>%
  mutate(
    Relevant_Breadth = case_when(
      Net_Predictive_Value > 0 ~ breadth_present,  # Positive = presence expected
      Net_Predictive_Value < 0 ~ breadth_absent,   # Negative = absence expected
      TRUE ~ NA_real_  # Zero/neutral case
    )
  ) %>%
  select(gene, mean_shap_pres, mean_shap_abs, breadth_present, breadth_absent, mean_abs_shap, Net_Predictive_Value, Relevant_Breadth, Env)

full_model_breadth_data %>% arrange(desc(mean_abs_shap)) %>%
    select(gene, mean_shap_pres, mean_shap_abs, breadth_present, breadth_absent, mean_abs_shap)

top_477_genes_full <- full_model_breadth_data %>% 
  arrange(desc(mean_abs_shap)) %>%
  distinct(gene, .keep_all = TRUE) %>%
  slice(1:477) %>%
  select(gene, mean_shap_pres, mean_shap_abs, breadth_present, breadth_absent)

top_477_genes_full_filtered <- full_model_breadth_data %>% 
    filter(gene %in% top_477_genes_full$gene) %>%
    arrange(desc(mean_abs_shap)) %>%
    select(gene, mean_shap_pres, mean_shap_abs, breadth_present, breadth_absent) %>%
    mutate(
      # Calculate Net Predictive Value
      Net_Predictive_Value = mean_shap_pres - mean_shap_abs,
      # Select appropriate breadth metric based on sign of NPV
      Relevant_Breadth = case_when(
        Net_Predictive_Value > 0 ~ breadth_present,  # Positive = presence expected
        Net_Predictive_Value < 0 ~ breadth_absent,   # Negative = absence expected
        TRUE ~ NA_real_  # Zero/neutral case
      )
    )

top_477_genes_full_filtered %>% select(Relevant_Breadth) %>% summary()

final_iterative_genes_breadth %>% select(Relevant_Breadth) %>% summary()

processed_genomovar_breadth <- final_iterative_genes_breadth %>%
  mutate(ModelType = "Iterative")

processed_genomovar_breadth_full <- top_477_genes_full_filtered %>%
  mutate(ModelType = "Full")

processed_genomovar_breadth_combined <- bind_rows(processed_genomovar_breadth, processed_genomovar_breadth_full)

saveRDS(processed_genomovar_breadth_combined, 
        file = data_here("00_data", "Figure_Processed_Data", "processed_genomovar_breadth_combined.RDS"))

```

## Assign variants as Mut or WT
```{r}
gene_info <- pa_gene_ids %>%
 mutate(Variant_Type = ifelse(PID == 100 & Query_Cov == 100, "WT", "Mut"))

saveRDS(gene_info, 
        file = data_here("00_data", "Figure_Processed_Data", "gene_variant_PA_Annotations.rds"))
```

# ============================================
# Figure 4
# ============================================

## Calculate Gene Variability
```{r}
### I'm pretty sure we don't use this, if so can be removed. CHECK THIS. 

# ---- Prepare Data for Variability Plot ----
niche_groups <- list(
  Chronic = c("Adult CF", "Pediatric CF", "Bronchiectasis"),
  Acute = c("Fecal", "Blood", "Urinary", "Wound")
)

variability_plot_data <- calculate_variability_plot_data(
  flat_df_annotated,
  niche_groups,
  min_niches_per_group = 1,
  require_consensus = "other" #changed from "all"
)

# ---- Join Gene Names to Variability Data ----
variability_plot_data <- variability_plot_data %>%
  left_join(
    flat_df_annotated_gene_names %>%
      select(gene, aa_cluster, Non.unique.Gene.name) %>%
      distinct(gene, .keep_all = TRUE),
    by = "gene"
  )

# ---- Save Variability Data ----
saveRDS(variability_plot_data, 
        file = data_here("00_data", "Figure_Processed_Data", "variability_plot_data.RDS"))

# ---- Identify "top" genes to label ----
genes_to_label <- variability_plot_data %>%
  arrange(desc(abs(NPV_difference))) %>%
  filter(abs(NPV_difference) >= 0.055) %>%
  distinct(gene, Group, .keep_all = TRUE) %>%
  select(gene, Group, Non.unique.Gene.name, Annotation_Inferred, NPV_difference, variability_focal)

genes_to_label %>% print(n = Inf)

# ---- Save Genes to Label ----
saveRDS(genes_to_label, 
        file = data_here("00_data", "Figure_Processed_Data", "genes_to_label_variability_plot.RDS"))
```

## Load Iris Gene Data
```{r}
### These need different names and the paths need to be updated to the new data location.

# Get all subdirectories (which are PA_IDs)
pa_id_dirs <- list.dirs(iris_output_dir, recursive = FALSE, full.names = TRUE)

# Load all _funct_no_funct_alt.csv files and add PA_ID column
iris_gene_funct_data <- purrr::map_dfr(pa_id_dirs, function(dir_path) {
  pa_id <- basename(dir_path)  # Extract PA ID from directory name
  file_path <- file.path(dir_path, paste0(pa_id, "_funct_no_funct_alt.csv"))
  
  if (file.exists(file_path)) {
    read.csv(file_path) %>%
      mutate(PA_ID = pa_id, .before = 1)
  } else {
    NULL
  }
}) %>%
  bind_rows()

saveRDS(iris_gene_funct_data, 
        file = data_here("00_data", "Figure_Processed_Data", "chronic_acute_gene_funct_data.RDS"))

# Get all subdirectories (which are PA_IDs) for the v3 data ("final" set of genes for plot)
pa_id_dirs_v3 <- list.dirs(iris_output_dir_v3, recursive = FALSE, full.names = TRUE)

# Load all _funct_no_funct_alt.csv files and add PA_ID column for v3 data
iris_gene_funct_data_v3 <- purrr::map_dfr(pa_id_dirs_v3, function(dir_path) {
  pa_id <- basename(dir_path)  # Extract PA ID from directory name
  file_path <- file.path(dir_path, paste0(pa_id, "_funct_no_funct_alt.csv"))

  if (file.exists(file_path)) {
    read.csv(file_path) %>%
      mutate(PA_ID = pa_id, .before = 1)
  } else {
    NULL
  }
}) %>%
  bind_rows()

saveRDS(iris_gene_funct_data_v3, 
        file = data_here("00_data", "Figure_Processed_Data", "chronic_acute_gene_funct_data_v3.RDS"))
```

# ============================================
# Figure 5
# ============================================

## Process Data for the Phylogenetic Tree
```{r}
# ---- Helper Function: Clean Names ----
clean_name <- function(x) {
  x %>%
    gsub(".fasta", "", .) %>%
    gsub(" = ", "_", .) %>%
    gsub("=", "_", .) %>%
    gsub("/|\\(.*\\)| grande colonie", "", .) %>%
    gsub("\\s+|-|\\.", "", .) %>%
    toupper()
}

# ---- Match Tree Tips to Metadata ----
tree_names_clean <- clean_name(Experimental_Tree$tip.label)
metadata_names_clean <- clean_name(Experimental_Metadata$Original.ID)
overlap <- intersect(tree_names_clean, metadata_names_clean)

cat("Number of overlapping tips/metadata:", length(overlap), "\n")

# Prune tree to overlapping tips
pruned_tree <- ape::drop.tip(Experimental_Tree, Experimental_Tree$tip.label[!tree_names_clean %in% overlap])

# ---- Assign Environment Categories ----
Experimental_Metadata <- Experimental_Metadata %>%
  filter(clean_name(Original.ID) %in% overlap)

# ---- Merge in Growth Curve Fits ----
Experimental_Metadata <- Experimental_Metadata %>%
  left_join(gc_fits, by = c("Library_ID" = "strain"))

Experimental_Metadata %>% select(Genome, Env, r) %>% print(n = Inf)

# ---- Match Tree to Metadata ----
Experimental_clean_name_metadata <- Experimental_Metadata %>%
  mutate(treeNames = clean_name(Original.ID))

# ---- Prune tree to final filtered metadata ----
filtered_tips_clean <- clean_name(Experimental_clean_name_metadata$Original.ID)

filtered_tree <- ape::drop.tip(
  pruned_tree, 
  pruned_tree$tip.label[!clean_name(pruned_tree$tip.label) %in% filtered_tips_clean]
)

# ---- Clean Names for Tree and Metadata ----
filtered_tree$tip.label <- clean_name(filtered_tree$tip.label)
Experimental_clean_name_metadata$treeNames <- clean_name(Experimental_clean_name_metadata$treeNames)

Experimental_clean_name_metadata %>% select(Genome, Env, r) %>% print(n = Inf)

# ---- Update Col names ----
Experimental_clean_name_metadata <- Experimental_clean_name_metadata %>%
  rename(
    "Elastase Production"    = ecr.n,
    "Biofilm Fraction"       = bfrac,
    "Growth Rate"            = r,
  )

# ---- Save Processed Tree ----
saveRDS(filtered_tree, 
        file = data_here("00_data", "Figure_Processed_Data", "Iterative_Approach", "filtered_Experimental_tree.RDS"))
```

## Process Data for PCA on Phenotype
```{r}
### I don't think I use the PCA results here anymore, probably remove, but check first. 

# ---- Phylogenetic Clustering ----
phylo_dist <- cophenetic(filtered_tree)
hc <- hclust(as.dist(phylo_dist), method = "average")
clusters <- cutree(hc, k = 5)

# Add cluster assignments to metadata
Experimental_clean_name_metadata$cluster <- clusters[base::match(
  clean_name(Experimental_clean_name_metadata$Original.ID), 
  clean_name(names(clusters))
)]

# ---- Perform PCA ----
Experimental_clean_name_metadata %>% colnames() %>% head(n = 90) 

cols <- c("Elastase Production", 
          "Biofilm Fraction",           
           "Growth Rate"
          )

pca_data <- Experimental_clean_name_metadata[, cols] %>% na.omit()
pca_res <- prcomp(pca_data, scale. = TRUE)

# ---- Save PCA Results and Metadata ----
saveRDS(pca_res, 
        file = data_here("00_data", "Figure_Processed_Data", "pca_res_Experimental_phenotypes.RDS"))  

saveRDS(Experimental_clean_name_metadata, 
        file = data_here("00_data", "Figure_Processed_Data", "filtered_Experimental_metadata.RDS"))
```

## Summarize Stats for Paper
```{r}
### This should probably go in a different script, `Fig4_Experimental_Phenotype_Stats.R` or similar

# Adjust if your metadata object has a different name
df <- filtered_metadata

# Standardize environment labels if needed
df <- df %>%
  mutate(Env = recode(Env, "Non-CF Human Infection" = "Acute Human Infection"))

# Use your actual column names here
traits <- c(
  "Biofilm Fraction",
  "Growth Rate",
  "Elastase Production"
)

# Make sure traits are numeric
df <- df %>%
  mutate(across(all_of(traits), as.numeric))

# Summary by environment
env_summary <- df %>%
  group_by(Env) %>%
  summarise(
    n = sum(complete.cases(across(all_of(traits)))),
    biofilm_mean = mean(`Biofilm Fraction`, na.rm = TRUE),
    biofilm_sd   = sd(`Biofilm Fraction`, na.rm = TRUE),
    growth_mean  = mean(`Growth Rate`, na.rm = TRUE),
    growth_sd    = sd(`Growth Rate`, na.rm = TRUE),
    elastase_mean = mean(`Elastase Production`, na.rm = TRUE),
    elastase_sd   = sd(`Elastase Production`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Env)

# Optional: summaries for the specific statements you quoted
cf_summary <- env_summary %>% filter(grepl("CF", Env))
noncf_summary <- env_summary %>% filter(!grepl("CF", Env))
aquatic_summary <- env_summary %>% filter(grepl("Aquatic", Env))

# Print tables
print(env_summary)
print(cf_summary)
print(noncf_summary)
print(aquatic_summary)

# If you want the exact phrasing helper:
env_summary %>%
  transmute(
    Env,
    biofilm_txt = sprintf("biofilm fraction (mean = %.3f, n = %d)", biofilm_mean, n),
    growth_txt = sprintf("growth rate (mean = %.3f, n = %d)", growth_mean, n),
    elastase_txt = sprintf("elastase activity (mean = %.3f, n = %d)", elastase_mean, n)
  ) %>%
  print(n = Inf)
```