---
title: "Make Supplemental Figures"
output: html_document
date: "2026-04-07"
---

# Description
- This code produces the plots for the supplemental figures in the CCS paper. 

# Load Setup and Figure Config 
```{r setup, include=FALSE}
library(install.load)

install_load("here")

source(here("00_R_configs", "config.R"))

source(here("00_R_configs", "00_setup.R"))

source(here("00_R_configs", "00_functions.R"))
``` 

# Load Data 
```{r}
# ---- Load Metadata ----
metadata_unmapped <- readRDS(data_here("00_data", "Figure_Processed_Data", "metadata_bronch_fixed.RDS")) %>%
  mutate(Genome_ID = str_replace_all(Genome_ID, "\\.", "_"))

write.csv(metadata_unmapped, data_here("00_data", "Figure_Processed_Data", "metadata_bronch_fixed.csv"), row.names = FALSE
)
# ---- Load Phylogenetic Tree ----
phylo_tree_30k <- read.tree(data_here("00_data", "Figure_Processed_Data", "parsnp.tree"))

# ---- Load ANI Data ----
ani_data <- data_here("00_data", "ANI_Clusters/synthetic_group") %>%
  list.files(pattern = "\\.csv$", full.names = TRUE) %>%
  purrr::map(~{
    df <- read_csv(.x)
    cluster <- str_extract(basename(.x), "\\d+\\.\\d+")
    colnames(df)[2] <- cluster
    df
  }) %>%
  reduce(full_join, by = colnames(.[[1]])[1])

# ---- Load Category Mappings ----
source(here("00_R_configs", "00_functional_category_mapping.R"))

# ---- Load Blocked Cross Performance Data ----
blocked_performance_data <- data_here("00_data", "Figure_Processed_Data", "Blocked_Cross_Results") %>%
  list.files(pattern = "_holdout_stats_newPanaroo\\.csv$", full.names = TRUE) %>%
  set_names(~str_extract(basename(.), "^[a-z]+") %>% str_to_title()) %>%
  purrr::map(read.csv) 
  # Loads everything in the folder, needs to be filtered downstream

# ---- Load "Full" Model Breadth ----
full_model_breadth_data <- readRDS(data_here("00_data", "Figure_Processed_Data", "full_model_breadth.rds"))

# ---- Load Filtered Metadata ----
metadata_filtered <- readRDS(data_here("00_data", "Figure_Processed_Data", "Metadata_Filtered.rds"))

probs_long <- readRDS(data_here("00_data", "Figure_Processed_Data", "477_probs_processed.rds"))

# ---- Data ----
#TODO: NEED TO MOVE RELEVANT BITS TO `06_Process_Data_For_Figures.RMD`

df_pao1 <- load_all_gene_files("00_data/Mutant-Calling-11K-CCS-Top27shap/results/ouputs_for_graphing") %>%
  mutate(reference = "PAO1", .after = 1)

flat_df_annotated_gene_names <- readRDS(
  file = data_here("00_data", "Figure_Processed_Data", "flat_df_annotated_gene_names_SHAP_data.RDS")
)

pa_gene_ids <- read.csv(
  data_here("00_data", "PAO1_blast_result_new.csv")) %>% as.tibble()

gene_info <- readRDS(
  file = data_here("00_data", "Figure_Processed_Data", "gene_variant_PA_Annotations.rds")
)

metadata_mapped <- readRDS(
  file = data_here("00_data", "Figure_Processed_Data", "metadata_mapped.rds")
)

cladeB_blocked_confusion_matrix_data <- read.csv(
  file = data_here("00_data", "Figure_Processed_Data", "test_cannonical_cladeB.csv")
)

metadata_mapped

stats_df_merged <- df_pao1 %>%
  left_join(metadata_mapped, by = c("Genome" = "Genome_ID")) %>%
  dplyr::filter(!is.na(Niche)) %>%
  filter(Clade == "CladeA") %>%
  left_join(pa_gene_ids, by = c("gene" = "Locus.Tag")) %>%
  mutate(
    Mut_Status = case_when(
      Mutation == "Contig split" ~ "Alternative function", # CALLING CONTIG SPLIT AS ALT FUNCTION FOR NOW.
      TRUE ~ as.character(Mut_Status)
    )
  )

stats_df_merged_noContigSplits <- df_pao1 %>%
  left_join(metadata_mapped, by = c("Genome" = "Genome_ID")) %>%
  dplyr::filter(!is.na(Niche)) %>%
  filter(Clade == "CladeA") %>%
  left_join(pa_gene_ids, by = c("gene" = "Locus.Tag")) %>%
  filter(Mutation != "Contig split")

# ---- Load Clade B Confusion Matrix Data ----
cladeB_blocked_confusion_matrix_data <- read.csv(data_here("00_data", "Figure_Processed_Data", "test_cannonical_cladeB.csv")) %>%
  as.tibble()
```

# Process Data
```{r}
# ---- Define Niche Mapping ----
niche_map <- c(
  "Bronchiectasis" = "Bronchiectasis",
  "CF" = "Adult CF",
  "early.CF" = "Pediatric CF",
  "Early CF" = "Pediatric CF",
  "Early.CF" = "Pediatric CF",
  "Urinary" = "Urinary",
  "Gastrointestinal" = "Gastrointestinal",
  "Wound" = "Wound",
  "Blood" = "Blood",
  "Aquatic" = "Aquatic",
  "Ocean" = "Aquatic",
  "Hospital Environment" = "Hospital",
  "Hospital.Environment" = "Hospital",
  "Hospital.Enviornment" = "Hospital",
  "Rectal/Feces" = "Fecal",
  "Rectal-Feces" = "Fecal",
  "Rectal.Feces" = "Fecal"
)

# ---- Apply to Metadata ----
metadata_all_clades <- metadata_unmapped %>%
  mutate(Niche = remap_niches(Niche)) %>%
  filter(!is.na(Niche))

# ---- Define environment order ----
env_labels <- c(
  "Aquatic", "Hospital",
  "Adult CF", "Pediatric CF", "Bronchiectasis", "Blood", "Urinary", "Fecal", "Wound"
)

# ---- Organize Niche Data for Confusion Matrix and Rename ----
cladeB_confusion_matrix_long <- cladeB_blocked_confusion_matrix_data %>%
  mutate(
    # Extract individual niche labels from the list strings
    actual_niches = str_extract_all(True.Labels, "(?<=')[^']+(?=')"),
    predicted_niches = str_extract_all(Predicted.Labels, "(?<=')[^']+(?=')")
  ) %>%
  # Unnest to expand lists into individual rows
  unnest(cols = c(actual_niches, predicted_niches), 
         names_sep = "_") %>%
  rename(
    Actual_Niche = actual_niches,
    Predicted_Niche = predicted_niches
  ) %>%
  # Remap niche names
  mutate(
    Actual_Niche = remap_niches(Actual_Niche),
    Predicted_Niche = remap_niches(Predicted_Niche)
  ) %>%
  # Filter out NAs from remapping
  filter(!is.na(Actual_Niche), !is.na(Predicted_Niche)) %>%
  # Now group and summarize
  group_by(Actual_Niche, Predicted_Niche) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  group_by(Actual_Niche) %>%
  mutate(
    row_pct = n / sum(n),
    row_total = sum(n)
  ) %>%
  ungroup()
```

# Make Plots

## Sup Fig 1: Singleton Counts per ANI Level
```{r}
# ---- Merge with Metadata ----
metadata_filtered_prepped <- metadata_filtered %>%
  mutate(Genome_ID = str_replace_all(Genome_ID, "\\.", "_"))

key_col <- colnames(ani_data)[1]
merged_data <- left_join(
  metadata_filtered_prepped, 
  ani_data, 
  by = c("Genome_ID" = key_col)
)

# ---- Reshape to Long Format ----
ani_long <- merged_data %>%
  pivot_longer(
    cols = matches("^\\d+\\.\\d+$"),
    names_to = "ANI_Cluster",
    values_to = "ANI_Value"
  ) %>%
  filter(!is.na(ANI_Value)) %>%
  add_count(ANI_Cluster, ANI_Value, name = "value_count") %>%
  select(-value_count)

# ---- Calculate Niche Diversity per ANI Group ----
ani_summary <- ani_long %>%
  filter(!is.na(ANI_Value), !is.na(Niche)) %>%
  group_by(ANI_Cluster, ANI_Value) %>%
  summarise(
    num_unique_niches = pmin(n_distinct(Niche), 9),
    group_size = n(),
    .groups = "drop"
  )

# ---- Filter to Selected ANI Thresholds ----
ani_keep <- c(99.33, 99.5, 99.75, 99.96, 99.99)
ani_summary_filtered <- ani_summary %>% 
  filter(ANI_Cluster %in% ani_keep)

# ---- Calculate Weighted Mean per ANI Threshold ----
ani_mean_points <- ani_summary_filtered %>%
  group_by(ANI_Cluster) %>%
  summarise(
    mean_niches = weighted.mean(num_unique_niches, group_size),
    .groups = "drop"
  )

# ---- Plot ANI vs. ENV Diversity ----
p.ani_env <- ggplot(
  ani_summary_filtered, 
  aes(x = ANI_Cluster, y = num_unique_niches)
) +
  geom_jitter(aes(color = group_size), width = 0.2, alpha = 0.8) +
  scale_color_gradientn(
    colors = c("#F3E8F0", "#CFA3C2", "#9A4F7F", "#914174", "#883268"),
    trans = "log10",
    limits = c(10, 1000),
    oob = scales::squish,
    breaks = c(10, 25, 50, 100, 500),
    labels = c("10", "25", "50", "100", "500"),
    name = "Samples in group"
  ) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  geom_point(
    data = ani_mean_points,
    aes(x = ANI_Cluster, y = mean_niches),
    color = "black", 
    shape = 8, 
    size = 2, 
    stroke = 1,
    inherit.aes = FALSE
  ) +
  scale_y_continuous(breaks = 1:9, limits = c(1, 9)) +
  labs(x = "ANI Cluster", y = "Number of Environments") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1, size = 8),
    axis.title.x = element_text(margin = margin(t = 15))
  )

ggsave(
  filename = data_here("00_data", "Figures", "SupFig1_ANIdepth.png"),
  plot = p.ani_env,
  width = 4,
  height = 3,
  dpi = 300
)
```

## Sup Fig 2: Full Tree (Clades A & B & C)
```{r}
#TODO: Fix random mis-labled tip in Clade B

#  ---- Clean Tip Labels ----
phylo_tree_30k$tip.label <- gsub(".fna$", "", phylo_tree_30k$tip.label)

# ---- Add Outgroup ----
outgroup_tip <- "PA7_GCA_000017205.1_ASM1720v1_genomic"

metadata_for_join <- metadata_all_clades %>% 
  select(Genome_ID, Clade, Niche, Genomovar_Cluster, Clone_Cluster) %>%
  filter(Clade %in% c("CladeA", "CladeB", "CladeC"))

# ---- Remove problematic samples (mislabeled tips) ----
samples_to_remove <- c("GCA_028683235_1", "GCA_003976125_1", "GCA_039570425_1", "GCA_003055425_1")
metadata_for_join <- metadata_for_join %>%
  filter(!(Genome_ID %in% samples_to_remove))

if (!(outgroup_tip %in% metadata_for_join$Genome_ID)) {
  outgroup_row <- data.frame(
    Genome_ID = outgroup_tip, 
    Clade = "Outgroup", 
    Niche = "Outgroup", 
    Genomovar_Cluster = "Outgroup", 
    Clone_Cluster = "Outgroup"
  )
  metadata_for_join <- rbind(metadata_for_join, outgroup_row)
}

metadata_for_join <- metadata_for_join %>% filter(Clade %in% c("CladeA", "CladeB", "CladeC", "Outgroup"))

# ---- Filter and Root Tree ----
phylo_rooted <- phylo_tree_30k %>%
  ape::keep.tip(phylo_tree_30k$tip.label[phylo_tree_30k$tip.label %in% metadata_for_join$Genome_ID]) %>%
  root(outgroup = outgroup_tip, resolve.root = TRUE)

# ---- Remove Outgroup from Tree ----
outgroup_tips <- as_tibble(phylo_rooted_treedata_bothClades) %>%
  filter(Clade == "Outgroup") %>%
  pull(label)

pruned_tree <- ape::drop.tip(phylo_rooted, outgroup_tips)

# ---- Join Metadata to Tree ----
phylo_rooted_treedata <- as.treedata(pruned_tree)

metadata_for_join_clade <- metadata_for_join %>%
  mutate(label = Genome_ID) %>%
  select(label, Clade, Genomovar_Cluster, Clone_Cluster) %>%
  filter(!is.na(Clade))

phylo_rooted_treedata_bothClades <- left_join(
  phylo_rooted_treedata,
  metadata_for_join_clade,
  by = "label"
)

metadata_for_join_clade$Clade %>% unique()

# ---- Plot Tree ----
p.tree_both_clades <- ggtree(phylo_rooted_treedata_bothClades, layout = "rectangular") +
  geom_tree(color = "grey65", linewidth = 0.3) +
  geom_tippoint(aes(color = Clade), size = 0.7, alpha = 0.9, na.rm = TRUE) +
  theme_tree() +
  theme(legend.position = "right") +
  scale_color_manual(
    values = c("CladeA" = "#5D93AE", "CladeB" = "#E36B2E", "CladeC" = "#883268", "Outgroup" = "black"),
    na.translate = FALSE
  ) +
  theme(legend.title = element_blank())

# ---- Save Plot ----
ggsave(filename = data_here("00_data", "Figures", "SupFig2_FullTree.png"), 
  plot = p.tree_both_clades, 
  width = 10, 
  height = 12)

# ---- Plot Genomovars ----

# Get top 30 genomovars by count
top30_genomovars <- metadata_for_join_clade %>%
  count(Genomovar_Cluster, sort = TRUE) %>%
  slice_head(n = 30) %>%
  pull(Genomovar_Cluster)

# Add a display column: top 30 keep their label, rest become "Other"
phylo_rooted_treedata_bothClades <- phylo_rooted_treedata_bothClades %>%
  mutate(Genomovar_Display = ifelse(Genomovar_Cluster %in% top30_genomovars, Genomovar_Cluster, "Other"))

# Build color vector: top 30 get distinct colors, "Other" gets grey
top30_colors <- setNames(
  scales::hue_pal()(30),
  top30_genomovars
)
top30_colors["Other"] <- "grey75"

# Build tree once to extract tip positions
tree_plot_data <- ggtree(phylo_rooted_treedata_bothClades, layout = "rectangular")$data

# One label tip per top-30 cluster: the tip closest to the cluster's vertical midpoint
cluster_label_tips <- tree_plot_data %>%
  filter(isTip, Genomovar_Cluster %in% top30_genomovars) %>%
  group_by(Genomovar_Cluster) %>%
  slice_min(abs(y - median(y)), n = 1, with_ties = FALSE) %>%
  ungroup()

p.tree_genomovar <- ggtree(phylo_rooted_treedata_bothClades, layout = "rectangular") +
  geom_tree(color = "grey65", linewidth = 0.3) +
  geom_tippoint(aes(color = Genomovar_Display), size = 0.7, alpha = 0.9, na.rm = TRUE) +
  scale_color_manual(values = top30_colors, breaks = top30_genomovars) +
  geom_text(
    data = cluster_label_tips,
    aes(x = x, y = y, label = Genomovar_Cluster),
    hjust = -0.15, size = 2.5,
    inherit.aes = FALSE
  ) +
  theme_tree2() +
  theme(legend.position = "none")

# ---- Save Plot ----
ggsave(filename = data_here("00_data", "Figures", "SupFig2_FullTree_Genomovars.png"), 
  plot = p.tree_genomovar, 
  width = 10, 
  height = 12)

##TODO: REMOVE THIS

# Tip-level clade labels currently used for coloring
tip_tbl <- as_tibble(phylo_rooted_treedata_bothClades)

tip_df <- tip_tbl %>%
  filter(!is.na(label)) %>%
  select(label, Clade)

# Get tree and pairwise patristic distances
tr <- as.phylo(phylo_rooted_treedata_bothClades)
dmat <- cophenetic(tr)

# For each tip, check agreement with k nearest neighbors
k <- 20
tip_outlier_score <- map_dfr(rownames(dmat), function(tip) {
  nn <- setdiff(rownames(dmat), tip)
  nn <- nn[order(dmat[tip, nn])][1:k]

  tip_clade <- tip_df$Clade[base::match(tip, tip_df$label)]
  nn_clade <- tip_df$Clade[base::match(nn, tip_df$label)]

  tibble(
    label = tip,
    Clade = tip_clade,
    same_clade_frac = mean(nn_clade == tip_clade, na.rm = TRUE),
    nn_majority_clade = names(sort(table(nn_clade), decreasing = TRUE))[1]
  )
}) %>%
  arrange(same_clade_frac)

# Most suspicious tips first
tip_outlier_score %>% print(n = 30)

# 1 GCA_028683235_1 CladeA            0.15 CladeB           
# 2 GCA_003976125_1 CladeA            0.15 CladeB           
# 3 GCA_039570425_1 CladeA            0.15 CladeB           
# 4 GCA_003055425_1 CladeA            0.15 CladeB   

# Print Clone Details for Clade B
clone_cluster_summary_B <- metadata_unmapped %>%
  filter(Clade == "CladeB") %>%
  group_by(Clone_Cluster) %>%
  summarise(
    member_count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(member_count))

clone_cluster_summary_A <- metadata_unmapped %>%
  filter(Clade == "CladeA") %>%
  group_by(Clone_Cluster) %>%
  summarise(
    member_count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(member_count))

# View the results
print(clone_cluster_summary_A, n = 30) 

# Calculate percentage of SAMPLES in clusters with > 20 members
sample_pct_over_20_B <- metadata_unmapped %>%
  filter(Clade == "CladeB") %>%
  left_join(
    clone_cluster_summary_B %>% 
      select(Clone_Cluster, member_count),
    by = "Clone_Cluster"
  ) %>%
  summarise(
    total_samples = n(),
    samples_in_clusters_over_20 = sum(member_count > 20),
    pct_samples_over_20 = (sum(member_count > 20) / n()) * 100
  )

sample_pct_over_20_A <- metadata_unmapped %>%
  filter(Clade == "CladeA") %>%
  left_join(
    clone_cluster_summary_A %>% 
      select(Clone_Cluster, member_count),
    by = "Clone_Cluster"
  ) %>%
  summarise(
    total_samples = n(),
    samples_in_clusters_over_20 = sum(member_count > 20),
    pct_samples_over_20 = (sum(member_count > 20) / n()) * 100
  )

```

## Sup Fig 3: Clade B Blocked Cross
```{r}
# ---- Complete Confusion Matrix with All Combinations (Including Zeros) ----
cladeB_cm_complete <- expand_grid(
  True = factor(env_labels, levels = env_labels),
  Pred = factor(env_labels, levels = env_labels)
) %>%
  left_join(
    cladeB_confusion_matrix_long %>%
      rename(True = Actual_Niche, Pred = Predicted_Niche) %>%
      mutate(
        True = factor(True, levels = env_labels),
        Pred = factor(Pred, levels = env_labels)
      ),
    by = c("True", "Pred")
  ) %>%
  mutate(
    n = replace_na(n, 0),
    row_pct = replace_na(row_pct, 0),
    row_total = replace_na(row_total, 0),
    True = factor(True, levels = env_labels),
    Pred = factor(Pred, levels = env_labels)
  )

# ---- Plot Confusion Matrix Heatmap ----
p.cladeB_confusion_matrix <- ggplot(cladeB_cm_complete, aes(x = Pred, y = True, fill = row_pct)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = n), size = 3.5, color = "black") +
  scale_fill_gradient(
    low = "white", 
    high = "#883268",
    limits = c(0, 1), 
    breaks = seq(0, 0.8, 0.1)
  ) +
  labs(
    x = "Predicted labels",
    y = "Actual labels",
    fill = "Proportion of\nRow Total"
  ) +
  theme_bw(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 10, margin = margin(t = 20)),
    axis.title.y = element_text(size = 10),
    legend.position = "none",
    panel.grid = element_blank()
  )

# ---- Calculate Per-Environment Accuracy Metrics ----
# Get total samples per environment
total_samples <- sum(cladeB_cm_complete$n)
class_totals <- cladeB_cm_complete %>%
  group_by(True) %>%
  summarise(class_n = sum(n), .groups = "drop") %>%
  mutate(null_accuracy = (class_n / total_samples) * 100)

cladeB_accuracy_df <- cladeB_cm_complete %>%
  group_by(True) %>%
  summarise(
    correct_n = sum(n * as.numeric(True == Pred)),
    total_n = sum(n),
    actual_accuracy = (correct_n / total_n) * 100,
    .groups = "drop"
  ) %>%
  rename(category = True) %>%
  left_join(
    class_totals %>% select(True, null_accuracy),
    by = c("category" = "True")
  ) %>%
  pivot_longer(
    cols = c(actual_accuracy, null_accuracy),
    names_to = "type",
    values_to = "accuracy"
  ) %>%
  mutate(type = ifelse(type == "actual_accuracy", "actual", "null")) %>%
  select(category, type, accuracy)

# ---- Plot Accuracy Comparison ----
accuracy_colors <- c(
  actual = "#cfa3c2", 
  null   = "#883268"
)

p.cladeB_accuracy <- ggplot(cladeB_accuracy_df, aes(x = category, y = accuracy, fill = type, alpha = type)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(
    values = accuracy_colors,
    labels = c(actual = "Actual Accuracy", null = "Null Accuracy"),
    name = "Accuracy Type"
  ) +
  scale_alpha_manual(values = c(actual = 0.5, null = 1), guide = "none") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.35))) +
  labs(x = NULL, y = "Accuracy (%)") +
  theme_bw(base_size = 18) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.position = "none"
  )

# ---- Combine Confusion Matrix and Accuracy Plots ----
p.cladeB_conf_mat_combined <- (p.cladeB_confusion_matrix | p.cladeB_accuracy) + 
  plot_layout(widths = c(2, 1))

# ---- Save Plot ----
ggsave(
  filename = data_here("00_data", "Figures", "SupFig3_ConfusionMatrix_CladeB.png"),
  plot = p.cladeB_conf_mat_combined,
  width = 12,
  height = 6,
  dpi = 300
)
```

## Sup Fig 4: Env Counts per Genomovar
```{r}
# ---- Filter to Top 50 Largest Clusters ----
niche_order <- names(niche_colors)

top_50_clusters <- metadata_filtered %>%
  count(Genomovar_Cluster, name = "cluster_size") %>%
  slice_max(cluster_size, n = 50) %>%
  pull(Genomovar_Cluster)

metadata_filtered_top50 <- metadata_filtered %>%
  filter(Genomovar_Cluster %in% top_50_clusters)

# ---- Calculate Niche Counts and Prepare Plot Data ----

counts <- metadata_filtered_top50 %>%
  count(Genomovar_Cluster, Niche, name = "count")

# Order clusters by total count (descending)
counts_plot <- counts %>%
  mutate(
    Genomovar_Cluster = fct_reorder(Genomovar_Cluster, count, .fun = sum, .desc = TRUE),
    Niche = factor(Niche, levels = niche_order)
  )

genomovar_order <- levels(counts_plot$Genomovar_Cluster)

# Create complete grid for heatmap (fills missing combinations with 0)
counts_complete <- counts %>%
  complete(Genomovar_Cluster, Niche, fill = list(count = 0)) %>%
  mutate(
    fill_var = if_else(count > 0, as.character(Niche), NA_character_),
    Niche = factor(Niche, levels = rev(niche_order)),
    Genomovar_Cluster = factor(Genomovar_Cluster, levels = genomovar_order)
  )



# ---- Top Panel: Stacked Bar Chart ----

p.genomovar_dist <- ggplot(counts_plot, aes(x = Genomovar_Cluster, y = count, fill = Niche)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = niche_colors, name = "Env") +
  theme_bw(base_size = 14) +
  labs(x = "Genomovar Cluster", y = "Env Abundance") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    #axis.ticks.x = element_blank(),
    #axis.title.x = element_blank(),
    axis.line.x = element_line(),
    plot.margin = margin(0, 0, 0, 0)#,
    #legend.position = "none"
  )

# ---- Bottom Panel: Presence/Absence Heatmap ----

p.niche_dist <- ggplot(counts_complete, aes(Genomovar_Cluster, Niche)) +
  geom_tile(aes(fill = fill_var), color = "white", width = 0.95, height = 0.95) +
  scale_fill_manual(values = niche_colors, na.value = "white", name = "Niche") +
  theme_bw(base_size = 14) +
  labs(x = NULL, y = "Env Presence") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.title.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  )

# ---- Create Legend ----

legend_grob <- cowplot::get_legend(
  ggplot(counts_plot, aes(x = Genomovar_Cluster, y = count, fill = Niche)) +
    geom_col() +
    scale_fill_manual(values = niche_colors, name = "Env") +
    theme_bw(base_size = 14) +
    theme(legend.position = "right")
)

# ---- Combine Panels ----

p.GenomovarDistribution <- p.genomovar_dist / p.niche_dist + plot_layout(heights = c(3, 2))

ggsave(
  filename = data_here("00_data", "Figures", "SupFig4_GenomovarDistribution.png"),
  plot = p.GenomovarDistribution,
  width = 15,
  height = 8,
  dpi = 300
)

ggsave(
  filename = data_here("00_data", "Figures", "SupFig4_GenomovarDistribution_Part1.png"),
  plot = p.genomovar_dist,
  width = 15,
  height = 8,
  dpi = 300
)
```

## Sup Fig 5: Model Performance per Env Grouping
```{r}
niche_colors_all = c(
  # === Level 1: Broad Categories ===
  "Host Associated"       = "#CB6650",
  "Environmental"         = "#72854E",
  
  # === Level 2: Primary Subgroups ===
  # Under Host Associated
  "Human"                 = "#D26749",
    "Other Clinical"                 = "#D26749",
  "Animal"                = "#B89C8D",
  
  # Under Environmental
  "Natural Environment"   = "#879667",
  "Human-Impacted Environment" = "#656565",
  
  # === Level 3: Secondary Subgroups ===
  # Under Human (Host Associated > Human)
  "Internal"              = "#B6502B",
  "External"              = "#C36557",
  "Respiratory"           = "#D4622D",
  "Other Internal"        = "#B6502B",
  
  # Under Natural Environment (Environmental > Natural)
  "Terrestrial"           = "#9CA780",
  "Aquatic"               = "#9CBED2",
  "Ocean"                 = "#5D93AE",
  
  # Under Human-Impacted Environment (Environmental > HumanImpacted)
  "Built Environment"     = "#8C8C8C",
  "Hospital"  = "#727272",
  "Waste Water"           = "#7F7F7F",
  
  # === Level 4: Specific Niches ===
  # Under Respiratory (Host Associated > Human > Respiratory)
  "Lung"                  = "#D4622D",
  "Other Lung"                  = "#D4622D",
  "Bronchiectasis"        = "#E36B2E",
  "Pneumonia"             = "#F28D32",
  "CF"                    = "#F3BA36",
  "Adult CF"                    = "#F3BA36",
  "Pediatric CF"              = "#F7D074",
  
  # Under Internal (Host Associated > Human > Internal)
  "Urinary"               = "#A8441B",
  "Gastrointestinal"      = "#944432",
  "Blood"                 = "#7F4348",
  "Rectal/Feces"          = "#4B2324",
  "Fecal"          = "#4B2324",
  
  # Under External (Host Associated > Human > External)
  "Wound"                 = "#C57066",
  "Burn"                  = "#CA9192",
  "Eye"                   = "#E3C5C5"
)


blocked_performance_data <- data_here("00_Panaroo_100_UCF", "Blocked_Cross_Results") %>%
  list.files(pattern = "_holdout_stats_newPanaroo\\.csv$", full.names = TRUE) %>%
  set_names(~str_extract(basename(.), "^[a-z]+") %>% str_to_title()) %>%
  purrr::map(read.csv) 
  # Loads everything in the folder, needs to be filtered downstream

# ---- Helper Function: Apply Niche Mapping to Label Strings ----
remap_labels <- function(label_string, niche_map) {
  result <- label_string
  for (old_name in names(niche_map)) {
    new_name <- niche_map[[old_name]]
    old_pattern <- paste0("'", old_name, "'")
    new_replacement <- paste0("'", new_name, "'")
    result <- gsub(old_pattern, new_replacement, result, fixed = TRUE)
  }
  return(result)
}

# ---- Apply to Performance Data (rename in True/Predicted Labels) ----
blocked_performance_data <- lapply(blocked_performance_data, function(df) {
  df %>%
    mutate(
      True.Labels = remap_labels(True.Labels, niche_map),
      Predicted.Labels = remap_labels(Predicted.Labels, niche_map)
    )
})

# ---- Filter: select which models to include ----
models_to_include <- c("First", "Eighth", "Ninth", "Tenth", "Eleventh", "Twelfth")

# ---- Parse label strings and compute per-niche accuracy ----
parse_labels <- function(label_string) {
  stringr::str_extract_all(label_string, "(?<=')[^']+(?=')")[[1]]
}

model_order <- c("First", "Eighth", "Ninth", "Tenth", "Eleventh", "Twelfth")

# rename map: old -> new
model_name_map <- c(
  "First"    = "Round 1",
  "Eighth"   = "Round 2",
  "Ninth"    = "Round 3",
  "Tenth"    = "Round 4",
  "Eleventh" = "Round 6",
  "Twelfth"  = "Round 5"
)

env_accuracy_all <- blocked_performance_data %>%
  .[intersect(model_order, names(.))] %>%
  purrr::imap_dfr(function(df, model_name) {
    df %>%
      calc_env_accuracy(blocked = TRUE) %>%
      calc_normalized_accuracy() %>%
      mutate(
        ENV = ifelse(ENV == "Gastrointestinal", "Rectal/Feces", ENV),
        ENV = ifelse(ENV == "Rectal/Feces", "Fecal", ENV),
        ENV = ifelse(ENV == "CF", "Adult CF", ENV),
        ENV = ifelse(ENV == "Early CF", "Pediatric CF", ENV),
        ENV = ifelse(ENV == "Hospital Environment", "Hospital", ENV),
        Model = model_name
      )
  }) %>%
  mutate(
    Model = factor(Model, levels = model_order),
    Model_Display = recode(as.character(Model), !!!model_name_map),
    ENV = factor(ENV, levels = names(niche_colors))
    #Model_Display = factor(Model_Display, levels = unname(model_name_map[model_order]))
  )

# ---- Plot ----
p.blocked_niche_accuracy <- ggplot(env_accuracy_all, aes(x = ENV, y = accuracy, fill = ENV)) +
  geom_col() +
  geom_text(aes(label = scales::percent(accuracy, accuracy = 1)),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = niche_colors_all) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.15), expand = c(0, 0)) +
  facet_wrap(~Model_Display, ncol = 3) +
  labs(x = NULL, y = "Per-Environment Normalized Accuracy") +
  theme_bw(base_size = 13) +
  theme(
    axis.text.x  = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 9),
    legend.position = "none",
    strip.text   = element_text(face = "bold", size = 11),
    panel.spacing = unit(1, "lines")
  )

p.blocked_niche_accuracy

ggsave(
  filename = data_here("00_data", "Figures", "SupFig5_Niche_Accuracy.png"),
  plot = p.blocked_niche_accuracy,
  width = 15,
  height = 8,
  dpi = 300
)

```

## Sup Fig 6: Per Env ML Model Probability Distributions
```{r}
# Identify focal environment per sample from REAL metadata
probs_summary <- probs_long %>%
  left_join(
    metadata_mapped %>% 
      select(Genome_ID, Niche) %>% 
      rename(Genome_IDs = Genome_ID, Focal_Env = Niche),
    by = "Genome_IDs"
  )

#---- Histogram of Probability Distributions by Focal_Env ----
probs_summary$Env <- factor(probs_summary$Env, levels = env_order)
probs_summary$Focal_Env <- factor(probs_summary$Focal_Env, levels = env_order)

p.prob_dist <- ggplot(probs_summary, aes(x = Probability, fill = Env)) +
  geom_histogram(position = "dodge", bins = 10, alpha = 1) +
  facet_wrap(~Focal_Env, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = niche_colors) +
  labs(
    x = "Probability",
    y = "Frequency",
    fill = "Environment",
    title = "Probability Distributions by Focal Environment"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    strip.text = element_text(size = 10, face = "bold")
  )

# ---- Save the plot ----
ggsave(
  filename = data_here("00_data", "Figures", "SupFig6_Prob_Distributions.png"),
  plot = p.prob_dist,
  width = 12,
  height = 10,
  dpi = 300
)
```

## Sup Fig 7: Loss of Function Mutations for Relevant Genes
```{r}
niche_colors <- c(
  "Aquatic"        = "#9CBED2",
  "Hospital"       = "#727272",
  "Adult CF"       = "#F3BA36",
  "Pediatric CF"   = "#F7D074",
  "Bronchiectasis" = "#E36B2E",
  "Blood"          = "#7F4348",
  "Urinary"        = "#A8441B",
  "Fecal"          = "#B89C8D",
  "Wound"          = "#C57066"
)
###TODO: These depend on data and functions from:
# /home/emehlferber3/02_Github/Env_XGBoost/00_R_configs/02_Make_Figures/Fig3_Update/Fig3_TopGenesStatistics.R
# Pull thos and put into main function script, etc. 

stats_df_merged_lof <- stats_df_merged %>%
  mutate(
    case_when(
      Mut_Status == "No function" ~ "No function",
      Mut_Status != "No function" ~ "Functional",
      TRUE ~ NA_character_
    ))

lof_glmer_results <- run_glmer_analysis(
  df_input = stats_df_merged_lof,
  reference_name = "PAO1",
  random_effect_var = c("Genomovar_Cluster"),
  niche_ref = "Adult CF",
  mut_wt_values = c("Functional"),
  mut_case_values = c("No function"),
  genes_to_analyze = top_genes
)

lof_cld_plot <- plot_mutation_probabilities_cld(
  glmer_results = lof_glmer_results,
  genes = top_genes,
  niche_colors = niche_colors,
  ncol = 3
)

ggsave(
  filename = data_here("00_data", "Figures", "SupFig7_LossOfFunction_MutationProbabilities_CLD.png"),
  plot = lof_cld_plot$plot,
  width = 7,
  height = 4,
  dpi = 300
)
```

## Sup Fig 8: Percentage Loss of Function per Gene per Env
```{r}
gene_mutation_summary <- stats_df_merged %>%
  filter(gene %in% top_genes) %>%
  group_by(gene, Niche, Gene.Name) %>%
  summarise(
    Total = n(),
    Mutated = sum(Mut_Status == "No function", na.rm = TRUE),
    Mutation_Rate = Mutated / Total,
    .groups = "drop"
  ) %>%
  mutate(
    gene = factor(gene, levels = top_genes),
    Niche = factor(Niche, levels = names(niche_colors)),
    Gene_Label = ifelse(is.na(Gene.Name) | Gene.Name == "", as.character(gene), as.character(Gene.Name))
  ) %>%
  arrange(gene, Niche) %>%
  mutate(Gene_Label = factor(Gene_Label, levels = unique(Gene_Label))) %>%
  as_tibble()

mutation_rate_plot <- ggplot(gene_mutation_summary, aes(x = Niche, y = Mutation_Rate, fill = Niche)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  facet_wrap(~ Gene_Label, ncol = 3) +
  scale_fill_manual(values = niche_colors, drop = FALSE) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = ggplot2::expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "Environment",
    y = "Observed loss-of-function rate"
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

ggsave(
  filename = data_here("00_data", "Figures", "SupFig8_PercentageLossOfFunction_PerGene_PerEnv.png"),
  plot = mutation_rate_plot,
  width = 7,
  height = 4,
  dpi = 300
)
```

## Sup Fig 9: Mutation probabilities with Contig Splits Removed
```{r}
glmer_results_noContigSplits <- run_glmer_analysis(
  df_input = stats_df_merged_noContigSplits,
  reference_name = "PAO1",
  random_effect_var = c("Genomovar_Cluster"),
  niche_ref = "Adult CF",
  mut_wt_values = c("Functional"),
  mut_case_values = NULL,
  genes_to_analyze = top_genes
)

noContigSplits_cld_plot <- plot_mutation_probabilities_cld(
  glmer_results = glmer_results_noContigSplits,
  genes = top_genes,
  niche_colors = niche_colors,
  ncol = 3
)

ggsave(
  filename = data_here("00_data", "Figures", "SupFig9_Mutation_Probabilities_NoContigSplits.png"),
  plot = noContigSplits_cld_plot$plot,
  width = 7,
  height = 4,
  dpi = 300
)
```


