---
title: "Make Supplemental Tables/CSV Files"
output: html_document
date: "2026-04-22"
---

# Description
- This code produces the Supplemental Tables for the CCS paper.     

# Load Setup and Figure Config 
```{r setup, include=FALSE}
library(install.load)

install_load("here")

source(here("config.R"))

source(here("Iterative_Model_Approach", "00_setup.R"))

source(here("Iterative_Model_Approach", "00_functions.R"))
``` 

# Load Data 
```{r}
flat_df_annotated_gene_names <- readRDS(
  file = data_here("00_data", "Figure_Processed_Data", "flat_df_annotated_gene_names_SHAP_data.RDS")
)

pa_gene_ids <- read.csv(
  data_here("00_data", "PAO1_blast_result_new.csv")) %>% as.tibble()

gene_info <- readRDS(
  file = data_here("00_data", "Figure_Processed_Data", "gene_variant_PA_Annotations.rds")
)
```

# Sup Table 1: Metadata for all 477 Genes (CSV file)
```{r}
# Check structure first
nrow(flat_df_annotated_gene_names)  # Total rows
n_distinct(flat_df_annotated_gene_names$gene)  # Unique genes

# Create wider format with Env-specific SHAP columns
# Keep only the specified metadata columns + SHAP values
wide_df_gene_data <- flat_df_annotated_gene_names %>%
  select(gene, Env, 
         mean_shap_pres, mean_shap_abs, mean_abs_shap,
         study_bias, aa_cluster, aa_length, longest_aa,
         Non.unique.Gene.name, Annotation, Annotation_Inferred,
         gene_name_cluster_inferred, Level1, Level2) %>%
  group_by(gene) %>%
  # Take first value of metadata columns per gene
  mutate(across(c(study_bias, aa_cluster, aa_length, longest_aa,
                  Non.unique.Gene.name, Annotation, Annotation_Inferred,
                  gene_name_cluster_inferred, Level1, Level2),
               first)) %>%
  ungroup() %>%
  # Now pivot only SHAP columns by Env
  pivot_wider(
    id_cols = c(gene, study_bias, aa_cluster, aa_length, longest_aa,
                Non.unique.Gene.name, Annotation, Annotation_Inferred,
                gene_name_cluster_inferred, Level1, Level2),
    names_from = "Env",
    values_from = c("mean_shap_pres", "mean_shap_abs", "mean_abs_shap"),
    names_sep = "_"
  )

# VERIFICATION
nrow(wide_df_gene_data)  # Should be 477
n_distinct(wide_df_gene_data$gene)  # Should be 477

# Rename columns and drop unwanted ones
wide_df_gene_data_named <- wide_df_gene_data %>%
  rename(
    Protein_Variant_ID = gene,
    Gene_Name = gene_name_cluster_inferred,
    Functional_Annotation = Annotation_Inferred,
    Protein_Length = aa_length,
    Protein_Cluster = aa_cluster,
    Longest_Protein_in_Cluster = longest_aa
  ) %>%
  # Rename SHAP columns for each environment
  rename_with(~str_replace(., "mean_shap_pres_", "Mean_SHAP_FeaturePresent_"), 
              starts_with("mean_shap_pres_")) %>%
  rename_with(~str_replace(., "mean_shap_abs_", "Mean_SHAP_FeatureAbsent_"), 
              starts_with("mean_shap_abs_")) %>%
  rename_with(~str_replace(., "mean_abs_shap_", "Mean_Absolute_SHAP_Value_"), 
              starts_with("mean_abs_shap_")) %>%
  # Keep only the columns we want
  select(Protein_Variant_ID, Gene_Name, Functional_Annotation, Protein_Length, 
         Protein_Cluster, Longest_Protein_in_Cluster,
         starts_with("Mean_SHAP_FeaturePresent_"),
         starts_with("Mean_SHAP_FeatureAbsent_"),
         starts_with("Mean_Absolute_SHAP_Value_"))

# Merge with gene_info
wide_df_gene_data_named_merged <- wide_df_gene_data_named %>%
  left_join(
    gene_info %>%
      select(Prot_acc, Locus.Tag, Variant_Type, Gene.Name, PID, Aln_length, Query_Cov) %>%
      rename(PA_Gene_ID = Locus.Tag),
    by = c("Protein_Variant_ID" = "Prot_acc")
  ) %>%
  # Reorder columns to put new columns right after Protein_Variant_ID
  select(Protein_Variant_ID, PA_Gene_ID, Variant_Type, 
         Gene.Name, Functional_Annotation, Protein_Length,
         Protein_Cluster, Longest_Protein_in_Cluster, 
         PID, Aln_length, Query_Cov,
         starts_with("Mean_SHAP_FeaturePresent_"),
         starts_with("Mean_SHAP_FeatureAbsent_"),
         starts_with("Mean_Absolute_SHAP_Value_")) %>%
    rename(PA_Gene_Name = Gene.Name) %>%
    rename(Alignment_Length = Aln_length) %>%
    rename(Query_Coverage = Query_Cov) %>%
    rename(Percent_Identity = PID) %>%
    # Replace empty strings and any missing values with NA
    mutate(across(everything(), ~ifelse(. == "" | is.na(.), NA, as.character(.)))) %>%
    # Tag genes as Defined or Hypothetical
    mutate(Gene_Type = if_else(
      str_starts(Protein_Variant_ID, "group") & is.na(PA_Gene_Name),
      "Hypothetical",
      "Defined"
    ))

# Verify
nrow(wide_df_gene_data_named_merged)

# Count NA values in PA_Gene_Name where Protein_Variant_ID starts with "group"
# Count unique PA_Gene_IDs (or count NA entries individually)
na_count_group_genes <- wide_df_gene_data_named_merged %>%
  filter(str_starts(Protein_Variant_ID, "group")) %>%
  filter(is.na(PA_Gene_Name)) %>%
  group_by(PA_Gene_ID, .add = FALSE) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(count = if_else(is.na(PA_Gene_ID), n, 1)) %>%
  summarise(total = sum(count)) %>%
  pull(total)

na_count_group_genes

# Count "group" genes in top 10% of ANY Mean_Absolute_SHAP_Value column
# Calculate 90th percentile for each Mean_Absolute_SHAP_Value column
group_genes_top10_percent <- wide_df_gene_data_named_merged %>%
  filter(str_starts(Protein_Variant_ID, "group")) %>%
  filter(is.na(PA_Gene_Name)) %>%
  select(Protein_Variant_ID, starts_with("Mean_Absolute_SHAP_Value_")) %>%
  pivot_longer(-Protein_Variant_ID, 
               names_to = "shap_column", 
               values_to = "shap_value") %>%
  mutate(shap_value = as.numeric(shap_value)) %>%
  left_join(
    wide_df_gene_data_named_merged %>%
      select(starts_with("Mean_Absolute_SHAP_Value_")) %>%
      pivot_longer(everything(),
                   names_to = "shap_column",
                   values_to = "value") %>%
      mutate(value = as.numeric(value)) %>%
      group_by(shap_column) %>%
      summarise(threshold = quantile(value, 0.9, na.rm = TRUE), .groups = "drop"),
    by = "shap_column"
  ) %>%
  filter(!is.na(shap_value) & shap_value >= threshold) %>%
  distinct(Protein_Variant_ID) %>%
  nrow()

group_genes_top10_percent

# SANITY CHECK: Show counts per environment
sanity_check <- wide_df_gene_data_named_merged %>%
  filter(str_starts(Protein_Variant_ID, "group")) %>%
  filter(is.na(PA_Gene_Name)) %>%
  select(Protein_Variant_ID, starts_with("Mean_Absolute_SHAP_Value_")) %>%
  pivot_longer(-Protein_Variant_ID, 
               names_to = "shap_column", 
               values_to = "shap_value") %>%
  mutate(shap_value = as.numeric(shap_value)) %>%
  left_join(
    wide_df_gene_data_named_merged %>%
      select(starts_with("Mean_Absolute_SHAP_Value_")) %>%
      pivot_longer(everything(),
                   names_to = "shap_column",
                   values_to = "value") %>%
      mutate(value = as.numeric(value)) %>%
      group_by(shap_column) %>%
      summarise(threshold = quantile(value, 0.9, na.rm = TRUE), 
                .groups = "drop"),
    by = "shap_column"
  ) %>%
  mutate(in_top10 = !is.na(shap_value) & shap_value >= threshold) %>%
  group_by(shap_column) %>%
  summarise(n_group_genes_in_top10 = sum(in_top10, na.rm = TRUE),
            .groups = "drop")

sanity_check

# Total group genes with NA PA_Gene_Name
total_group_genes_na <- wide_df_gene_data_named_merged %>%
  filter(str_starts(Protein_Variant_ID, "group")) %>%
  filter(is.na(PA_Gene_Name)) %>%
  distinct(Protein_Variant_ID) %>%
  nrow()

total_group_genes_na

# Save as CSV
write.csv(
  wide_df_gene_data_named_merged %>%
    mutate(Avg_Mean_Absolute_SHAP = rowMeans(
      sapply(select(., starts_with("Mean_Absolute_SHAP_Value_")), as.numeric),
      na.rm = TRUE
    )) %>%
    arrange(desc(Avg_Mean_Absolute_SHAP)) %>%
    select(-Avg_Mean_Absolute_SHAP) %>%
    relocate(Gene_Type, .after = Protein_Variant_ID),
  file = data_here("00_data", "Figures", "Sup_Table_1_Gene_Metadata.csv"),
  row.names = FALSE
)

pa_gene_ids %>% filter(Locus.Tag %in% c("PA2591"))

gene_info %>% filter(Locus.Tag %in% c("PA2591"))
```

## Sup Table 4: Stats for Experimental Strains (CSV file)
```{r}
filtered_experimental_metadata <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "filtered_experimental_metadata.RDS")
)

# Adjust if your metadata object has a different name
df <- filtered_experimental_metadata

# Standardize environment labels if needed
df <- df %>%
  mutate(Env = recode(Env, "Non-CF Human Infection" = "Acute Human Infection"))

# Use your actual column names here
traits <- c(
  "Biofilm Fraction",
  "Growth Rate",
  "Elastase Production"
)

# Make sure traits are numeric
df <- df %>%
  mutate(across(all_of(traits), as.numeric))

# Exact-n descriptive statistics by environment and trait.
# Each row represents the complete-case sample size used for that trait within that environment.
stats_long <- df %>%
  select(Env, all_of(traits)) %>%
  pivot_longer(
    cols = all_of(traits),
    names_to = "Trait",
    values_to = "Value"
  ) %>%
  group_by(Env, Trait) %>%
  summarise(
    n_exact = sum(!is.na(Value)),
    mean = mean(Value, na.rm = TRUE),
    sd = sd(Value, na.rm = TRUE),
    se = ifelse(n_exact > 0, sd / sqrt(n_exact), NA_real_),
    ci_low = ifelse(n_exact > 1, mean + qt(0.025, df = n_exact - 1) * se, NA_real_),
    ci_high = ifelse(n_exact > 1, mean + qt(0.975, df = n_exact - 1) * se, NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    Trait = factor(Trait, levels = traits),
    Env = factor(Env)
  ) %>%
  arrange(Trait, Env)

# Wide summary for easy manuscript/table use.
env_summary <- stats_long %>%
  select(Env, Trait, n_exact, mean, sd, se, ci_low, ci_high) %>%
  pivot_wider(
    names_from = Trait,
    values_from = c(n_exact, mean, sd, se, ci_low, ci_high),
    names_sep = "__"
  ) %>%
  arrange(Env)

# Methods metadata for Nature-style reporting.
stats_methods_summary <- tibble::tibble(
  Analysis = c(
    "Descriptive summaries by environment",
    "Error bars",
    "Replicate definition",
    "Exact n reporting"
  ),
  Details = c(
    "Mean, SD, SE, and 95% CI computed separately for each trait within each environment",
    "95% confidence intervals from the t distribution (two-sided)",
    "Each genome / isolate is treated as one independent replicate",
    "Exact complete-case n is reported for each trait/environment combination"
  )
)

# Optional convenience tables for manuscript wording.
cf_summary <- stats_long %>% filter(grepl("CF", Env))
noncf_summary <- stats_long %>% filter(!grepl("CF", Env))
aquatic_summary <- stats_long %>% filter(grepl("Aquatic", Env))

# Exact replicate counts per environment (one row per Env).
env_n_summary <- df %>%
  group_by(Env) %>%
  summarise(n_env = n(), .groups = "drop") %>%
  arrange(Env)

# Print concise, publication-useful tables.
print(env_n_summary, n = Inf)
print(stats_long, n = Inf)
print(env_summary, n = Inf)
print(stats_methods_summary, n = Inf)

# Export tables as CSV files.
stats_export_dir <- "00_data/Figure_Processed_Data"
dir.create(stats_export_dir, recursive = TRUE, showWarnings = FALSE)
write_csv(env_n_summary, file.path(stats_export_dir, "paper_stats_n_per_env.csv"))
write_csv(stats_long, file.path(stats_export_dir, "paper_stats_long_summary.csv"))
write_csv(env_summary, file.path(stats_export_dir, "paper_stats_wide_summary.csv"))
write_csv(stats_methods_summary, file.path(stats_export_dir, "paper_stats_methods_summary.csv"))

# Keep the helper output if you want plain-text manuscript phrasing.
stats_long %>%
  transmute(
    Env,
    Trait,
    summary_txt = sprintf("%s: mean = %.3f, SD = %.3f, n = %d", Trait, mean, sd, n_exact)
  ) %>%
  print(n = Inf)
```

### I Need to figure out where I'm saving all the stats for Data S_4

## Sup Tables: Fig3 Top Genes Statistics

### Load exported model statistics objects and export as CSVs (separate for each analysis)
```{r}
library(here)

# Define output directory for supplementary table CSVs
output_dir <- data_here("00_data", "Figure_Processed_Data")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# ---- All Data Analysis ----
model_details_all_data <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "model_details_All_Data.RDS")
)
n_per_group_all_data <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "n_per_group_All_Data.RDS")
)
pairwise_stats_all_data <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "pairwise_stats_All_Data.RDS")
)

write.csv(
  model_details_all_data,
  file = file.path(output_dir, "SupTable_Model_Details_All_Data.csv"),
  row.names = FALSE
)
write.csv(
  n_per_group_all_data,
  file = file.path(output_dir, "SupTable_N_Per_Group_All_Data.csv"),
  row.names = FALSE
)
write.csv(
  pairwise_stats_all_data,
  file = file.path(output_dir, "SupTable_Pairwise_Stats_All_Data.csv"),
  row.names = FALSE
)

# ---- No Contig Splits Analysis ----
model_details_no_contig <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "model_details_No_Contig_Splits.RDS")
)
n_per_group_no_contig <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "n_per_group_No_Contig_Splits.RDS")
)
pairwise_stats_no_contig <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "pairwise_stats_No_Contig_Splits.RDS")
)

write.csv(
  model_details_no_contig,
  file = file.path(output_dir, "SupTable_Model_Details_No_Contig_Splits.csv"),
  row.names = FALSE
)
write.csv(
  n_per_group_no_contig,
  file = file.path(output_dir, "SupTable_N_Per_Group_No_Contig_Splits.csv"),
  row.names = FALSE
)
write.csv(
  pairwise_stats_no_contig,
  file = file.path(output_dir, "SupTable_Pairwise_Stats_No_Contig_Splits.csv"),
  row.names = FALSE
)

# ---- Loss-of-Function Analysis ----
model_details_lof <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "model_details_Loss_Of_Function.RDS")
)
n_per_group_lof <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "n_per_group_Loss_Of_Function.RDS")
)
pairwise_stats_lof <- readRDS(
  data_here("00_data", "Figure_Processed_Data", "pairwise_stats_Loss_Of_Function.RDS")
)

write.csv(
  model_details_lof,
  file = file.path(output_dir, "SupTable_Model_Details_Loss_Of_Function.csv"),
  row.names = FALSE
)
write.csv(
  n_per_group_lof,
  file = file.path(output_dir, "SupTable_N_Per_Group_Loss_Of_Function.csv"),
  row.names = FALSE
)
write.csv(
  pairwise_stats_lof,
  file = file.path(output_dir, "SupTable_Pairwise_Stats_Loss_Of_Function.csv"),
  row.names = FALSE
)

# ---- Print summaries for verification ----
cat("\n=== All Data Analysis ===\n")
cat("Model Details:", nrow(model_details_all_data), "rows\n")
cat("N Per Group:", nrow(n_per_group_all_data), "rows\n")
cat("Pairwise Stats:", nrow(pairwise_stats_all_data), "rows\n")

cat("\n=== No Contig Splits Analysis ===\n")
cat("Model Details:", nrow(model_details_no_contig), "rows\n")
cat("N Per Group:", nrow(n_per_group_no_contig), "rows\n")
cat("Pairwise Stats:", nrow(pairwise_stats_no_contig), "rows\n")

cat("\n=== Loss-of-Function Analysis ===\n")
cat("Model Details:", nrow(model_details_lof), "rows\n")
cat("N Per Group:", nrow(n_per_group_lof), "rows\n")
cat("Pairwise Stats:", nrow(pairwise_stats_lof), "rows\n")

cat("\nCSV files have been saved to:\n")
cat(output_dir, "\n")
```